---
title: "Data 607 Import XML, JSON, HTML"
author: "Steve Phillips"
date: "2023-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
##XML,xml12, rjson, and jsonlite will all be used to parse our documents
library(xml2)
library(tidyverse)
library(XML)
library(rjson)
library(jsonlite)
library(readr)
library(RCurl)
```
## The xml file was difficult to parse cleanly as a dataframe, so I sliced each column from the parsed XML and formed a dataframe from this.

```{r}

data_xml = read_xml("https://raw.githubusercontent.com/sphill12/DATA607/main/books.xml")
xml_book <- xml_text(xml_find_all(data_xml, ".//Book_Name"))
xml_author <- xml_text(xml_find_all(data_xml, ".//Authors"))
xml_atr_1 <- xml_text(xml_find_all(data_xml, ".//Attribute_One"))
xml_atr_2 <- xml_text(xml_find_all(data_xml, ".//Attribute_Two"))
df_xml <- tibble(book = xml_book, author = xml_author, attribute_1 = xml_atr_1, attribute_2 = xml_atr_2)
df_xml = df_xml[-1,]
```

```{r}
df_xml
```


## The html document was much cleaner compared to the xml, and only required that the column names be added to the document.
```{r}
data_html <- getURL("https://raw.githubusercontent.com/sphill12/DATA607/main/books.html")
df_html <- readHTMLTable(data_html, as_html = TRUE, as.data.frame = TRUE)
df_html <- as.data.frame(df_html)
colnames(df_html) <- c("Book Name", "Authors", "Attribute One", "Attribute Two")
df_html
```
## The json document also cleanly was made into a dataframe. Once again, only column names were added before the dataset was satisfactory.
```{r}
data_json <- jsonlite::fromJSON("https://raw.githubusercontent.com/sphill12/DATA607/main/books.json")
df_json <- as.data.frame(data_json)
```
```{r}
colnames(df_json) <- c("Book Name", "Authors", "Attribute One", "Attribute Two")
df_json
```







